Who we are affects how we vote

### 2019-06-08 response_to_review.Rmd

---
title: "Response to Reviewer Feedback"
author: "Jeremy Forbes"
date: "08/06/2019"
output: html_document
---

## Editor's comments

*1. Address reviewer 1's question about why spatial imputation is required given that the census is reported at a low level of aggregation (lines 150-151) that could presumably be clumped to give a good approximation to electorate boundaries.*

...

*2. I note that you do mention the ecological fallacy (lines 285-286) but reviewer 1's reaction suggests that you should expand your remarks here.*

...

*3. It appears that your approach to variable selection has confused everybody. My understanding of what you have done is that:*
*- You begin with 65 variables (are they all the available census data or a subset?).*

...

*- Then you standardise the variables.*

*- Then you remove some variables to do with Age Bands and Christianity (how many variables out of the 65 does this leave? Were these variables totally removed or lumped in some way? Christianity still appears in table 2 so they obviously weren't all removed) *

...

*- Then you use the PCA to identify variables that covary. If variables appear together in a PC with loading higher than 0.15 they are combined together into a factor, e.g. a new variable Income replaces two previously existing variables. (In the description of this it is unclear how many variables are being combined in each factor,  the 'ands' and commas make things tricky, .e.g  is 'renting and government housing' one variable or two? Do the factors use the loading weights or just the signs?)*

...

*- The steps above get you from 65 variables down to 30 (I can count 10 variables that  get accounted for via the creation of factors, but that doesn't get from 65 to 30 so presumably some other variables have been removed? Or is this difference all in the age bands and variants of Christianity? More clarity is required)*

...

*4. As you note in the paper, it seems likely that there are still issues with multi-collinearity. Figs 6 and 7 show the variables with significant effects, do the results change much if you fit a model with just these 13 variables instead of the 30?*

...

## Associate Editor's comments
*Some conceptually simple imputation methods are used to align polling and census data to investigate associations between census variables and voting behaviour at electoral division level.*

...

*The main thing I found confusing about this paper was the variable selection procedure. It seems to be heading in the direction of four PCs, then it veers into six factors before ending with 30 variables. This all requires greater clarity and justification.*

...

*The authors opt for “visualization” of the fixed effects rather than the usual table of estimated parameters. Perhaps this is a good choice given that there are so many. But then spatial parameters are exhibited in a huge table. This seems inconsistent.*

...

*I found the writing unclear in a few places. What are the “time-varying restrictions” on a variable (line228)? And the description of a partial residual (line 276) doesn’t make sense to me.*

...

*Some of the interpretations of results seem a little naïve. For example, does Figure 4 really justify concluding that “electorates effectively voted independently” except for 2001 and 2016? (Is it a coincidence that the spatial autoregression is greatest in 2001 and 2016 where there is no need for imputation?)*

...

## Reviewer 1's comments

*Its goal is to account for – though the word ‘explain’ is used! – patterns of voting at Australian federal elections, using ecological data, but with no reference to the ecological fallacy.*

...

*The paper is written in virtual total ignorance of the large amount of work done on Australian voting patterns; apart from reference to a few papers by geographers there is not a single mention of work by political scientists – of which there is a great deal. As a consequence there is no theoretical background whatsoever, let alone hypotheses to be tested. It is banal empiricism: throw a lot of data (what are called ‘electoral socio-demographic variables’) at something and see what comes out.*

...

*The banality of this approach is illustrated by the finding that ‘more de facto relationships’ are associated with greater support for Labor. No rationale for this is offered – why should people in such relationships be more likely to vote Labor? And while there may be a statistically significant relationship (i.e. slope coefficient) it may be substantively trivial – a fact hidden because the data are transformed. If there is a small SD then the ‘real’ difference between a place with many de facto relationships and another with few will be small and will not account for much of the absolute variation in Labour support at all. Further, since all of the relationships with individual variables are partial, it may well be that when ‘all other variables are held constant’ the difference in Labour support is trivial.*

...

*Much is done – unnecessarily in my view – in preparing the data through imputation. This has two components. First there is the creation of census data for the 150 constituencies because boundaries of the large census reporting areas may not coincide with the constituencies’ at the time of an election. Imputation is deployed by assuming that if 50% of area A is in district X then 50% of A’s population is allocated to X without any recognition that area A may be internally heterogeneous. But that is in any case unnecessary. Australian census data are reported – as the authors note – for very small areas with populations of only a few hundred each, and there are shape files associated with these areas. It is thus a straightforward task to amalgamate the small areas into the electoral districts and get the ‘real’ data for each without any (or at best a very, very small amount of) imputation.*

...

*The second component involves imputing the values for each Census variable at the date of the election, which can never be more than three years – at most – distant from the nearest Census (either before or after the election). This seems totally unnecessary; the likelihood of any substantial change in the relative position of any district on any variable sufficient to impact on the regression outcomes is so minimal that a lot of effort could have been saved. It just wasn’t worth doing.*

...

*The authors recognise that there is likely to be collinearity among 65 Census variables – why those; we are never told?  And so, as others do, they did a pca of the data matrix, reducing it to four components – we are not told what rotation was employed to achieve simple structure. We are not given the results of this analysis, merely a naming of the six factors identified – how do you get six ‘factors’ from four components? That naming was based on the loadings and, we are told, that any loading greater than 0.15 (presumably either positive or negative) was ‘considered large’. Since loadings are the equivalent of correlations then a loading of 0.15 means that the component accounts for just 2.25 per cent of the variation in that variable. This is – quite frankly – ludicrous.*

...

*My expectation was that they would then fit a model using the four components as the independent variables, thus avoiding any collinearity and confounding. But they didn’t. Instead they selected 30 of the 65 variables – which must still be collinear; the resulting equation in Table 2 clearly implies this given the large number of (a) insignificant relations (normal with collinearity – see a recent paper in Quality and Quantity), (b) unreasonably large coefficients (-8.07 for de facto, for example), and (c) very large standard errors for many of the variables.*

...

*The authors fit a spatial autoregressive model, on the assumption that neighbouring places are more likely to be similar than distant places (the ‘first law of geography’). This is only a significant finding at two of the elections and is wrongly interpreted: it does not necessarily show that then ‘the political climate of an electorate appears to be affected by the attitudes of its neighbours’ but, much more likely, that neighbouring places share characteristics in common that affect voting behaviour but are not captured by the independent variables included in the model. Spatial autocorrelation may reflect spatial diffusion processes – but in which direction? – but is much more likely to reflect an under-specified model.*

...

*Having fitted the model we than get two pages of ‘interpretation’, which is inductivism of the worst kind – we’ve found some relationships, how do we account for them?*

...

## Reviewer 2's comments

*1. It seems very strange to assume that population is uniform within electorates (line 133) so that weighting is proportional to area, and not to the number of people in an area of intersection. An electorate on the urban fringe may be strongly affected by loss of a relatively small but highly populated area but little affected by incorporating a large rural area. A simple indirect way to get at this might be to address the question: Would the results change much if A_{s,t} was log area, for each s,t ?*

...

*2. It seems to me that the use of Principal Components analysis is unclear.*
*a. PC appears to have been used as a model selection criterion, with the variables related to the first four PCs included into the model in Table 2. However typically when principal components are computed they only involve the explanatory variables, not the response, and so there is no guarantee that all the first four PCs (and only them) will be related to the response. For example using the eechidna data for 2016 I found the fifth PC was extremely significant. But this is a minor point.*

...

*b. The variables which load highly on a PC are likely to be highly correlated, and so if these high-loading variables are used together in the regression model we may again have multicollinearity. I found this was the case when applying the author’s regression model to the 2016 data. This may be why several of the predictors in Table 2 are never related significantly related to the response in any of the years.*

...

*c. Conversely there could be a variable which is independent of the other variables (so representing 1 / 65th of the explanatory information) but by dint of that fact not be included in the first four PCs. In short, the narrative leaves me unconvinced that all the variables included are necessary or that all the relevant variables are included in the regression.*

...

*d. I presume the “factors” referred to lines 210-220 were created as the first principal component of the constituent raw variables?*

...

*e. The description of the variables included in “Education” is not clear.*

...

*f. I could not get the variable ‘OtherLanguageHome’ in 2016 to be significant, but it also had a high VIF with other explanatory variables. Please check this.*

...

*3. I am not convinced it is reasonable to summarise the effect of age by median age. Perhaps age effects are more at the edges of the age distribution than in the middle? I would have thought the percentage of the population in different age bands e.g. 20-34 years and Over 65years, might be significant. For 2016 the % aged 20-34 years was highly significant and removed median age.*

...

*4. Is there any tendency for the vote in an election to depend on how long the incumbent has been in power? In your cross-sectional models this would be confounded with any other effects in the intercept, but a comment would be nice to know. It would help in terms of interpreting the variations in intercept, and setting a reader’s mind to rest that the issue has been considered. Perhaps a something like a simple regression of average vote (or the intercept) on length of incumbency could address the issue.*

...

### 2019-06-08 anzjs_elections.Rmd

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremyforbes1995@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

@@

hume %>%
  kable(longtable=FALSE, booktabs=TRUE, digits=4,
        format='latex',escape=FALSE,
        col.names=c(
          "Source zone (2016)",
          "Proportion",
          "Source zone population",
          "Population allocated to purple region: $\\hat{P}_{s,t}$",
          "AusCitizen (\\%): $D_s$"),
    caption = "Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.") %>%
  kableExtra::kable_styling(font_size=10, latex_options='scale_down')
```

>>>>

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author: Jeremy Forbes$^*$, Dianne Cook, Rob J Hyndman
toc: false
documentclass: anzsauth
classoption: times, doublespace
fontfamily: times
bibliography: references.bib
biblio-style: anzsj
output: 
  bookdown::pdf_document2:
    fig_caption: yes
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: anzjs_preamble.tex
---

# Summary{-}

We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN).

**Keywords:** federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna

@@

hume %>%
  kable(longtable=FALSE, booktabs=TRUE, digits=c(0,4,0,0,1),
        format='latex',escape=FALSE,
        col.names=c(
          "Source zone (2016)",
          "Proportion",
          "Source zone population",
          "Population allocated to purple region: $\\hat{P}_{s,t}$",
          "AusCitizen (\\%): $D_s$"),
    caption = "Population from each intersecting Census source zone (2016) that is allocated to the target zone (purple region - Hume electoral boundary in 2013), and the corresponding proportion of Australian citizens in each of these source zones.") %>%
  kableExtra::kable_styling(font_size=12, latex_options='scale_down')
```

### 2019-06-08 elections.Rmd

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremyforbes1995@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna"
wpnumber:
jelcodes: C31, C33, D72
blind: false
cover: false
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

>>>>

---
title: "Spatial modelling of the two-party preferred vote in Australian federal elections: 2001--2016"
author:
- familyname: Forbes
  othernames: Jeremy
  address: Monash University\newline Melbourne, Australia
  email: jeremyforbes1995@gmail.com
  correspondingauthor: true
- familyname: Cook
  othernames: Dianne
  address: Monash University\newline Melbourne, Australia
- familyname: Hyndman
  othernames: Rob J
  address: Monash University\newline Melbourne, Australia
abstract: "We examine the relationships between electoral socio-demographic characteristics and two-party preference in the six Australian federal elections held between 2001 to 2016. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, spatio-temporal imputation is employed to estimate Census data for the electorates at the time of each election. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can be thought of as the unobserved political climate in each electorate. Over time, the impact of most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in each of the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect was insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the `eechidna` `R` package (available on CRAN)."
keywords: "federal election, Census, Australia, spatial modelling, imputation, data science, socio-demographics, electorates, R, eechidna"
wpnumber: 08/19
jelcodes: C31, C33, D72
blind: false
cover: true
toc: false
bibliography: references.bib
biblio-style: authoryear-comp
header-includes: \usepackage{float}
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

### 2019-06-08 elections_conversation.Rmd

A simple way to measure voting patterns is to consider the two-party preferred (2PP) vote, which is based on the tally of preferences for the Labor and Liberal candidates, ignoring all other candidates. By convention, this is recorded as a percentage preference in favour of the Liberal party --- for example, a 2PP value of 45% indicates that 45% of voters ranked the Liberal candidate higher than the Labor candidate, while the remaining 55% ranked them in the reverse order.

@@

Each election was modelled separately, to allow us to see any changes over time, and to account for changing electorate boundaries. In this article, we highlight the variables with the strongest relationship to the two-party preferred vote, or which have had substantial changes over time. The full analysis is available at https://robjhyndman.com/publications/elections/.

@@

This shows that the baseline of party preference has varied over the elections, with the biggest swing occurring in the 2007 election where the mean electorate shifted more than five percentage points in favour of the Labor party. The dots represent the estimated 2PP value, and the lines indicate a 95% confidence interval providing a guide to the uncertainty in the estimate. Where the vertical lines corresponding to each election cross the 50% horizontal line, the 2PP vote for an average electorate is not statistically distinguishable from 50% for that election.

@@

This graph shows the effect of median age on the 2PP vote. The vertical scale can be interpreted as a score measuring how much the 2PP vote differs between an average electorate, and an electorate with median age in the top 10% of all electorates (with all other socio-demographic variables unchanged). The reverse is true for young electorates --- the score can be interpreted as how much the 2PP vote would be reduced if an average electorate kept all socio-demographic variables fixed but median age was reduced to be in the bottom 10% of all eletorates.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appear more likely to support the Liberal party, as the `Incomes` factor has a positive effect on Liberal preference. The effect also seems to have been increasing over time.

@@

```{r resids, include = FALSE}
all_resids <- data.frame(
  Residuals = c(glsmod16$actual_residuals, glsmod13$actual_residuals, glsmod10$actual_residuals, glsmod07$actual_residuals, glsmod04$actual_residuals, glsmod01$actual_residuals),
  gls_residuals = c(glsmod16$residuals, glsmod13$residuals, glsmod10$residuals, glsmod07$residuals, glsmod04$residuals, glsmod01$residuals),
  bind_rows(
    glsmod16$my_data %>% left_join(tpp16 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod13$my_data %>% left_join(tpp13 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod10$my_data %>% left_join(tpp10 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod07$my_data %>% left_join(tpp07 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod04$my_data %>% left_join(tpp04 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod01$my_data %>% left_join(tpp01 %>% dplyr::select(DivisionNm, StateAb), by = "DivisionNm")
  )
)
# %>% dplyr::select(Residuals, DivisionNm, DivisionID, StateAb, year, LNP_Percent, gls_residuals)
```

It is often suggested that states have systematic differences that cause their electorates to vote differently. We can explore this effect by looking at the difference between the actual 2PP vote in each electorate and what our modelling would predict based only on electorate characteristics. We found a bias toward Labor in the ACT, and to a smaller extent in Tasmania, and a bias toward the Liberals in the Northern Territory. No other states showed a significantly different voting pattern from what you would expect using only socio-demographic information. That suggests that any differences between Queensland and Victoria (for example) are due to different voter socio-demographics.

```{r resstate, fig.cap = "Boxplot of residuals by state with jittered points. States ordered by median residual. A state-specific bias not captured by the model is evident.", fig.asp=0.6, out.width="70%", fig.pos = 'h', out.extra = '', eval=FALSE}
all_resids %>%
  ggplot(aes(x = fct_reorder(StateAb, Residuals), y = Residuals)) +
  geom_hline(aes(yintercept = 0), col = "grey50") +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, shape = 1) +
  guides(col = FALSE, shape = FALSE, DivisionNm = FALSE) +
  theme_bw() +
  labs(x = "State", y = "Residuals")
```

@@

```{r readcooksd, eval = FALSE}
read_chunk('cooksd.R')
```

```{r getcooksd, eval = FALSE}
<<GetCookDistance>>
```

```{r cooksd, eval = FALSE}
cooksd_df <- data.frame(
  Residuals = c(glsmod16$actual_residuals, glsmod13$actual_residuals, glsmod10$actual_residuals, glsmod07$actual_residuals, glsmod04$actual_residuals, glsmod01$actual_residuals),
  gls_residuals = c(glsmod16$residuals, glsmod13$residuals, glsmod10$residuals, glsmod07$residuals, glsmod04$residuals, glsmod01$residuals),
  cooksd = c(glsmod16$cooksd, glsmod13$cooksd, glsmod10$cooksd, glsmod07$cooksd, glsmod04$cooksd, glsmod01$cooksd),
  bind_rows(
    glsmod16$my_data %>% left_join(tpp16 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod13$my_data %>% left_join(tpp13 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod10$my_data %>% left_join(tpp10 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod07$my_data %>% left_join(tpp07 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod04$my_data %>% left_join(tpp04 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"),
    glsmod01$my_data %>% left_join(tpp01 %>% dplyr::select(DivisionNm, DivisionID, StateAb), by = "DivisionNm"))
)
```

```{r plotcooksd, fig.cap = "Electorates' Cook's Distance values in each election.", out.height = "15%", eval = FALSE}
cooksd_df %>%
  ggplot(aes(x = cooksd)) +
  geom_dotplot() +
  facet_wrap(~year, nrow = 1) +
  # coord_fixed(ratio = .5) +
  labs(x = "Cook's Distance", y = "") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 8), axis.text.y = element_text(size = 8))
```

```{r, eval = FALSE}
cooksd_df %>%
  filter(cooksd > 0.1) %>%
  select(DivisionNm, StateAb, year, cooksd, Residuals, LNP_Percent, everything()) %>%
  View()
```

Based on the distribution of the Cook's distance values, a Cook's distance greater than $0.1$ is considered to be influential and a potential outlier. The electorate of Sydney (NSW) has a large Cook's distance from 2001 to 2013, due to its diverse population (language, birthplace and religion), high number of defacto relationships, high income, high household mobility and small amount of workers in extractive and transformative jobs. It has remained a strong supporter of the Labor party and the Liberal vote is severely overpredicted by the model, making it an outlier. Nearby in metropolitan NSW, the electorate of Wentworth is found to be an outlier in all but the 2007 election. Although historically Liberal, its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. Liberal support in Wentworth is underpredicted by the model in each year, and more so with Turnbull as Liberal leader.

Lingiari, an electorate taking up almost all of the Northern Territory, is an outlier in the 2001--2007 elections due to its large Indigenous population, young age profile and low rates of property ownership. Fowler (NSW) has a diverse population with a high proportion of migrants, many Buddhists and Muslims, and has strong Labor support, making it influential in 2001, 2004 and 2010. Other electorates with large Cook's distance are Barton (NSW) and Leichhardt (QLD) in 2016, and Canberra (ACT) in 2007.

```{r, eval = FALSE}
x <- model_df %>%
  filter(year == "2016") %>%
  select(-c(year, DivisionNm)) %>%
  as.matrix()
myhat <- hat(x)

plot(
  glsmod07$cooksd,
  hat(
    model_df %>%
      filter(year == "2007") %>%
      select(-c(year, DivisionNm)) %>%
      as.matrix()
  )
)
```

# Conclusion

This paper explores the effects of electoral socio-demographic characteristics on the two-party preferred vote in the 2001--2016 elections, using information from the corresponding Australian federal elections and Censuses. As a Census does not always occur in the same year as an election, Census data for the 2004--2013 elections are generated by employing a method of spatio-temporal imputation. This imputes electoral socio-demographics for the electoral boundaries in place at the time of the election --- an approach that is distinctly different from previous work on modelling election outcomes, where Census and election data are typically joined without addressing their temporal differences. Before estimating a model, these socio-demographic variables are standardized (to adjust for changing variable scales) and many variables (representing similar information) are combined into factors, resulting in a reduced predictor set. A spatial error model is then estimated for each election, accounting for the inherent spatial structure of the data.

Across the past six elections, most of the socio-demographics that drive the electoral two-party preferred vote are found to remain steady, whilst a few (typically weaker) effects vary over time. Industry and type of work are particularly influential, with energy-related and manufacturing/construction jobs, as well as administrative roles being strongly linked with the Liberal party in all elections. Incomes have a similarly consistent effect, with higher income areas supporting Liberal. Higher levels of unemployment shift from weak association with Labor to a significant Liberal effect over the years, and higher education levels are associated with Labor from 2007 (although significant only in 2016). It is also found that electorates with higher household mobility support Liberal, birthplace diversity favours Labor and more de facto relationships align with Labor preference --- although marriages, family and household sizes have no material influence. Furthermore, the neighbourhood (spatial) effects are found to be positive in all elections, although significant only in 2001 and 2016, meaning that in the 2004--2013 elections, electorates effectively voted independently.

The findings in this paper complement the existing literature by modelling temporal trends, which as far as the authors are aware, has not been done previously for Australian elections using a regression framework. It is also the first study to model any Australian election since 2010 using Census information.

Additionally, a key contribution of this research is the wrangling of the raw data and imputed data sets for the 2004, 2007, 2010 and 2013 elections, which have been contributed to the `eechidna` `R` package --- providing a rich, accessible data resource for future Australian electoral analysis.

@@

This paper was produced using `RMarkdown` [@rmarkdown] and `knitr` [@knitr]. All corresponding code for this paper can be found in the github repository [github.com/jforbes14/eechidna-paper](https://github.com/jforbes14/eechidna-paper), and the data used is available in the `eechidna` package [@eechidna]. All raw data was obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government.

# Software

All election and Census datasets, along with electoral maps and more, are available in the `eechidna` (Exploring Election and Census Highly Informative Data Nationally for Australia) `R` package, which can be downloaded from CRAN. The `eechidna`  package makes it easy to look at the data from the Australian Federal elections and Censuses that occurred between 2001 and 2016. This study contributed a large revision to the `eechidna` package, which included the addition of election and Census data for 2001--2010, voting outcomes for polling booths and imputed Census data for election years. For more details on using `eechidna`, please see the articles (vignettes) on the github page [ropenscilabs.github.io/eechidna/](https://ropenscilabs.github.io/eechidna/).

The authors would like to sincerely thank Anthony Ebert, Heike Hofmann, Thomas Lumley, Ben Marwick, Carson Sievert, Mingzhu Sun, Dilini Talagala, Nicholas Tierney, Nathaniel Tomasetti, Earo Wang and Fang Zhou, all of whom have contributed to the `eechidna` package.

>>>>

A simple way to measure voting patterns is to consider the two-party preferred (2PP) vote, which is based on the tally of preferences for the Labor and Coalition candidates, ignoring all other candidates. We express this as a percentage preference in favour of the Coalition --- for example, a 2PP value of 45% indicates that 45% of voters ranked the Coalition candidate higher than the Labor candidate, while the remaining 55% ranked them in the reverse order.

@@

Each election was modelled separately, to allow us to see any changes over time, and to account for changing electorate boundaries. In this article, we highlight the variables with the strongest relationship to the two-party preferred vote, or which have had substantial changes over time.

@@

This shows that the baseline of party preference has varied over the elections, with the biggest swing occurring in the 2007 election where the average electorate shifted more than five percentage points in favour of the Labor party. The dots represent the estimated 2PP value, and the lines indicate a 95% confidence interval providing a guide to the uncertainty in the estimate. Where the vertical lines corresponding to each election cross the 50% horizontal line, the 2PP vote for an average electorate is not statistically distinguishable from 50% for that election.

@@

This graph shows the effect of median age on the 2PP vote. The vertical scale can be interpreted as a score measuring how much the 2PP vote differs between an average electorate, and an electorate with median age in the top 10% of all electorates (with all other socio-demographic variables unchanged). The reverse is true for young electorates --- the score can be interpreted as how much the 2PP vote would be reduced if an average electorate kept all socio-demographic variables fixed but median age was reduced to be in the bottom 10% of all electorates.

When the uncertainty range covers the horizontal 0 line, the resulting impact can be considered statistically undetectable. Here age is a statistically significant effect only in 2004 and 2007.

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies that higher income electorates appear more likely to support the Coalition, as the Incomes factor has a positive effect on Coalition preference. The effect also seems to have been increasing over time.

@@

It is often suggested that states have systematic differences that cause their electorates to vote differently. We can explore this effect by looking at the difference between the actual 2PP vote in each electorate and what our modelling would predict based only on electorate characteristics. We found a bias toward Labor in the ACT, and to a smaller extent in Tasmania, and a bias toward the Coalition in the Northern Territory. No other states showed a significantly different voting pattern from what you would expect using only socio-demographic information. That suggests that any differences between Queensland and Victoria (for example) are due to different voter socio-demographics.

@@

```{r outliers}
res <- bind_rows(
  glsmod16$my_data %>% mutate(pred=glsmod16$predictions),
  glsmod13$my_data %>% mutate(pred=glsmod13$predictions),
  glsmod10$my_data %>% mutate(pred=glsmod10$predictions),
  glsmod07$my_data %>% mutate(pred=glsmod07$predictions),
  glsmod04$my_data %>% mutate(pred=glsmod04$predictions),
  glsmod01$my_data %>% mutate(pred=glsmod01$predictions)
) %>%
  dplyr::select(DivisionNm, year, LNP_Percent, pred) %>%
  mutate(residuals = LNP_Percent - pred) %>%
  arrange(-abs(residuals)) %>%
  mutate(
    DivisionNm = str_to_title(DivisionNm)
  )
```

By this measure, the top five most surprising results in the last six elections were the following:

```{r surprises}
res %>% dplyr::select(DivisionNm, year, LNP_Percent, pred) %>% head(5) %>%
  kable(booktabs=TRUE, digits=c(0,0,1,1),
        col.names=c("Electorate", "Election", "Actual 2PP", "Predicted 2PP"))
```

Capricornia is on the Queensland coast, taking in Rockhampton, Yeppoon and parts of Mackay. It was won in 2001 and 2004 by Labour MP Kirsten Livermore, defying the socio-demographic prediction which suggested it would be a safe Coalition seat. She also won it in 2007 and 2010, but with a smaller number of votes, so the result was not so different from predicted. It is currently held by Michelle Landry for the Coalition.

Herbert is another Queensland seat taking in Townsville. It was won in 2013 for the Coalition by Ewen Jones, and in 2016 by Cathy O'Toole for Labor by just 37 votes. The model predicts a much lower vote for the Coalition than tends to happen in practice.

Wentworth probably needs no explanation --- the seat formerly held by Malcolm Turnbull which he won with a big majority in 2013. Its two-party vote jumped by over 10 percentage points in 2010 without experiencing any notable changes in its socio-demographic makeup --- implying that this may be the direct effect of its Liberal member, Malcolm Turnbull, becoming the leader of the Liberal party. Liberal support in Wentworth is underpredicted by the model in each year, and more so with Turnbull as Liberal leader.

# True Blue Seats

Another way the model can be used is to identify the seats with the strongest propensity to vote one way or the other, based on their socio-demographics.

The top five most blue-ribbon Coalition seats in the last six elections are listed below. Mitchell takes in the north-western outer suburbs of Sydney, including Baulkham Hills, Kellyville, and parts of Castle Hill. Maranoa occupies a large part of south west Queensland including towns such as Charleville, Kingaroy and Winton. Mallee is, of course, the north west corner of Victoria including Mildura, Swan Hill and Stawell.

```{r trueblue}
res %>% dplyr::select(DivisionNm, year, LNP_Percent, pred) %>% 
  arrange(-pred) %>%
  head(5) %>%
  kable(booktabs=TRUE, digits=c(0,0,1,1),
        col.names=c("Electorate", "Election", "Actual 2PP", "Predicted 2PP"))
```

The top five seats most likely to vote Labour, based on their socio-demographics, are listed below. Batman has since been abolished, but occupied the inner north of Melbourne. Gellibrand takes in the inner western suburbs of Melbourne including Altona and Williamstown, while Fowler is in the western suburbs of Sydney, including Cabramatta and Liverpool.

```{r truered}
res %>% dplyr::select(DivisionNm, year, LNP_Percent, pred) %>% 
  arrange(pred) %>%
  head(5) %>%
  kable(booktabs=TRUE, digits=c(0,0,1,1),
        col.names=c("Electorate", "Election", "Actual 2PP", "Predicted 2PP"))
```

@@

This analysis is based on a research paper coauthored by Rob Hyndman and Di Cook from the Department of Econometrics and Business Statistics at the Monash Business School, and Jeremy Forbes, a former Monash University Honours student in econometrics. The full analysis is available at https://robjhyndman.com/publications/elections/.

The code used for the analysis can be found in the github repository [github.com/jforbes14/eechidna-paper](https://github.com/jforbes14/eechidna-paper).

All election and Census data since 2001, along with electoral maps and more, are available in the [`eechidna`](https://cran.r-project.org/package=eechidna) (Exploring Election and Census Highly Informative Data Nationally for Australia) `R` package, which can be downloaded from CRAN.  The eechidna package uses data obtained from the Australian Electoral Commission, the Australian Bureau of Statistics and the Australian Government.
